---
output:
  html_document: default
  pdf_document: default
---
Redwine by David Havera
========================================================


```{r global_options, include=FALSE}
knitr::opts_chunk$set(echo=FALSE, warning=FALSE, message=FALSE)

# Load all of the packages that you end up using in your analysis in this code
# chunk.

# Notice that the parameter "echo" was set to FALSE for this code chunk. This
# prevents the code from displaying in the knitted HTML output. You should set
# echo=FALSE for all code chunks in your file, unless it makes sense for your
# report to show the code that generated a particular plot.

# The other parameters for "message" and "warning" should also be set to FALSE
# for other code chunks once you have verified that each plot comes out as you
# want it to. This will clean up the flow of your report.

library('ggplot2') # visualization
library('ggthemes') # visualization
library('scales') # visualization
library('gridExtra') # visualisation
library('corrplot') # visualisation
library('ggfortify') # visualisation
library('igraph') # visualisation
library('dplyr') # data manipulation
library('readr') # data input
library('tidyr') # data wrangling
library('stringr') # string manipulation
library('forcats') # factor manipulation
library('tidytext') # text mining
library('SnowballC') # text analysis
library('wordcloud') # test visualisation
library('GGally')
library('e1071')
library('randomForest')
library('caret')
library('mlbench')
library('ggfortify')
library('reshape')
```

```{r echo=FALSE, Load_the_Data}
# Load the Data
redwine <- read.csv("wineQualityReds.csv")
```


I will be evaluating a red wine quality dataset.  This dataset contains over 
1,599 red wines with 11 features consisting of the chemical properties of the 
wine.  At least 3 wine experts rated the wine on a scale from 0 (bad) to 10 
(excellent).  I will be analyzing this data set to determine what chemical 
properties influence the quality of red wine.


```{r echo=FALSE}
names(redwine)

```
```{r echo=FALSE}
str(redwine)
```
```{r echo=FALSE}
summary(redwine)
```

```{r echo=FALSE}
sum(is.na(redwine))
```
```{r}
redwine$X <- NULL
```


### What is the structure of your dataset?
There are 13 variables with 1,399 observations.  All variables are numeric with 
the exception of X and quality being integers.  There are no #N/A's in the data 
set.  I removed X from the dataset.  I will now explore each variable with a 
density plot.

```{r}
# create histograms for each attribute
create_plot <- function(variable) {
  return(ggplot(aes_string(x = variable), data = redwine) + 
           geom_density())}

create_plot('fixed.acidity')

```

Fixed acidity contains most acids involved with wine or fixed or nonvolatile that 
do not evaporate readily.  It is slightly positively skewed and has a mean of 8.3.

```{r}
create_plot('volatile.acidity')

```

Volatile acidity is the amount of acetic acid in wine, 
which at too high of levels can lead to an unpleasant, vinegar taste.  It is
slightly positively skewed and has a mean of 0.53.

```{r}
create_plot('citric.acid')

```

Citric acid is found in small quantities, citric acid can add 'freshness' 
and flavor to wines.  Citric acid has a mean of 0.27 and is positively skewed.

```{r}
create_plot('residual.sugar')

```

Residual sugar is the amount of sugar remaining after fermentation stops, 
it's rare to find wines with less than 1 gram/liter and wines with greater than
45 grams/liter are considered sweet.
Residual sugar has a mean of 2.5 and is positively skewed.

```{r}
create_plot('chlorides')

```

Chlorides is the amount of salt in the wine.
Chlorides has a mean of 0.09 and is positively skewed.

```{r}
create_plot('free.sulfur.dioxide')

```

Free sulfur dioxide is the free form of SO2 exists in equilibrium between 
molecular SO2 (as a dissolved gas) and bisulfite ion; it prevents microbial growth 
and the oxidation of wine.  
Free sulfur dioxide has a mean of 15.9 and is positively skewed.  

```{r}
create_plot('total.sulfur.dioxide')

```

Total sulfur dioxide is the amount of free and bound forms of S02; 
in low concentrations, SO2 is mostly undetectable in wine, 
but at free SO2 concentrations over 50 ppm, SO2 becomes evident in the nose 
and taste of wine.  Total sulfur dioxide has a mean of 46.5 and is positively 
skewed.  


```{r}
create_plot('density')

```

The density of water is close to that of water depending on the percent alcohol 
and sugar content. Density has a mean of 1 and is normally distributed.  

```{r}
create_plot('pH')


```

pH describes how acidic or basic a wine is on a scale from 0 (very acidic) to 14
(very basic).  pH appears to be normally distributed with a mean of 3.311.

```{r}
create_plot('sulphates')

```

Sulphates is a wine additive which can contribute to sulfur dioxide gas (S02) 
levels, wich acts as an antimicrobial and antioxidant.
It has a mean of 0.7 and is positively skewed.  

```{r}
create_plot('alcohol')
```
Alcohol is the % alcohol content of the wine.  It has a mean of 0.7 and is 
positively skewed.

I created histograms for all the features and it looks like several of the variables
are positively skewed.  I will dig deeper into this once I select the features 
of interest but it's good to know this set could benefit from log scaling.

```{r echo=FALSE}
res <- cor(redwine)
round(res, 2)

```


### What is/are the main feature(s) of interest in your dataset?
I decided on the following features of interest based on the highest correlation 
to quality: alcohol, volatile acidity, and sulphates.

```{r echo=FALSE}
 
a1 <- ggplot(redwine, aes(x = alcohol)) +
  geom_histogram(binwidth = 0.01) +
  labs(y = 'Alcohol %', x = 'Count',
          title = "Alcohol") 


a2 <- ggplot(redwine, aes(y = alcohol, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = alcohol), alpha = 0.3) +
  labs(y = 'Alcohol %', x = '',
          title = "Alcohol distribution") +
    coord_flip()
grid.arrange(a1, a2)

ggplot(aes(factor(quality), 
            alcohol), 
        data = redwine) +
  geom_jitter( alpha = .3)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)




```

Alcohol has a positive correlation to quality of 0.48 
and over 75% of the wines have at least 11% alcohol. It appears anything above 
13 and below 8 would be an outlier.

```{r echo=FALSE}


b1 <- ggplot(redwine, aes(x = sulphates)) +
  geom_histogram(binwidth = 0.01) +
  labs(y = 'Sulphates (g/L)', x = 'Count',
          title = "Sulphates") 


b2 <- ggplot(redwine, aes(y = sulphates, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = sulphates), alpha = 0.3) +
  labs(y = 'Sulphates (g/L)', x = '',
          title = "Sulphates distribution") +
  coord_flip()

grid.arrange(b1, b2)

 ggplot(aes(factor(quality), 
            sulphates), 
        data = redwine) +
  geom_jitter( alpha = .3)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)

```

Sulphates have a .25 correlation to quality and over 75% of the wines have at 
least a 0.75 level of sulphates.  It appears anything above 1 and below 0.25 
would be an outlier.

```{r echo=FALSE}

p1 <- ggplot(redwine, aes(x = volatile.acidity)) +
  geom_histogram(binwidth = 0.01) +
  labs(y = 'Volatile Acidity (g/L)', x = 'Count',
          title = "Volatile acidity distribution") 

p2 <- ggplot(redwine, aes(y = volatile.acidity, x = 1)) +
  geom_boxplot(color = 'blue') +
  geom_jitter(aes(x = 1, y = volatile.acidity), alpha = 0.3) +
  labs(y = 'Volatile Acidity (g/L)', x = '',
          title = "Volatile acidity distribution") +
  coord_flip()

grid.arrange(p1, p2)

 ggplot(aes(factor(quality), 
            volatile.acidity), 
        data = redwine) +
  geom_jitter( alpha = .3)  + 
  geom_boxplot( alpha = .5,color = 'blue')+ 
  stat_summary(fun.y = "mean", 
               geom = "point", 
               color = "red", 
               shape = 8, 
               size = 4)


```

Volatile acidity has a (.39) correlation to quality and over 75% of the wines 
have at least a 0.6 level of acidity.  It appears anything above 1 and 
below 0.1 would be an outlier.

### What other features in the dataset do you think will help support your \
I would like to further explore density, residual sugar,
fixed.acidity and citric.acid.


```{r echo=FALSE}
redwine$total.acidity <- redwine$volatile.acidity + redwine$fixed.acidity
```

```{r echo=FALSE}
cor(redwine$total.acidity, redwine$quality)
```

### Did you create any new variables from existing variables in the dataset?
I decided that fixed and voliatile acidity can be combined into total acidity. 
However, I discovered that total_acidity only had a correlation to 
quality of 0.085.

```{r echo=FALSE}

```
```{r echo=FALSE}
skewness(redwine$alcohol)
skewness(redwine$sulphates)
skewness(redwine$volatile.acidity)
skewness(redwine$quality)
```

### Of the features you investigated, were there any unusual distributions? \
Did you perform any operations on the data to tidy, adjust, or change the form \
of the data? If so, why did you do this?

All the features that I am interested are positively skewed with sulphates being
the most promently skewed.  Therefore, I decided to log transform these features
and we can see the improved skewness below.

```{r echo=FALSE}
redwine$quality_log <- log10(redwine$quality + 1)
redwine$volatile.acidity_log <- log10(redwine$volatile.acidity + 1)
redwine$alcohol_log <- log10(redwine$alcohol + 1)
redwine$sulphates_log <- log10(redwine$sulphates + 1)
redwine$total.acidity_log <- log10(redwine$total.acidity + 1)
```

```{r echo=FALSE}
skewness(redwine$alcohol_log)
skewness(redwine$sulphates_log)
skewness(redwine$volatile.acidity_log)
skewness(redwine$quality_log)
```

```{r}


```
We can see the log transform reduced the sknewess across the feature set.


```{r echo=FALSE}
#R2
alcohol.lm = lm(quality ~ alcohol, data=redwine)
summary(alcohol.lm)$r.squared

sulphates.lm = lm(quality ~ sulphates, data=redwine)
summary(sulphates.lm)$r.squared

volatile.acidity.lm = lm(quality ~ volatile.acidity, data=redwine)
summary(volatile.acidity.lm)$r.squared

```

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

I chose the variables with the higest correlation to quality: alcohol, 
volatile.acidity, and sulphates.  I ran the R2 for these features against 
quality and came up with a total R2 of 0.44, which means that 44% of quality 
can be explained by these variables.  
Sulphates looks particularly low at only 6%.


```{r echo=FALSE}
  ggplot(aes(x = fixed.acidity, 
           y = citric.acid   ), 
       data = redwine) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)
```


### Did you observe any interesting relationships between the other features \
(not the main feature(s) of interest)?

I found it interesting that fixed acidity had a correlation of 0.67 
to citric acid.  There seems to be a lot of features related to acid in the 
dataset.  If I am not able to accurately predict from the three features that 
I chose, then I will look into engineering another feature around acidity that 
combines citric acid.


### What was the strongest relationship you found?

The strongest relationship was 0.48 correlation of quality to alcohol.  
I expect that this is the most impactful feature to predict the quality of wine,
but we will later test this hypothesis.

### Talk about some of the relationships you observed in this part of the \
investigation. Were there features that strengthened each other in terms of \
looking at your feature(s) of interest?

Per the graph below, density has a strong negative correlation to alcohol of 
(0.5) and a positive correlation of 0.36 to residual sugars.  
This makes sense since the data set description of this feature says 
that density is dependent on alcohol levels and sugars.  
However, it's interesting to see the relationship 
in terms of correlation numbers. 

```{r}
 ggplot(aes(x = alcohol, 
           y = density  , color = factor(quality)), 
       data = redwine) +
      geom_point(alpha = 0.8, size = 2) +
      geom_smooth(method = "lm", se = FALSE,size=1)  +
  scale_color_brewer(type='seq',
                   guide=guide_legend(title='Quality'))
```


 


### Were there any interesting or surprising interactions between features?
```{r echo=FALSE}
  ggplot(aes(x = residual.sugar, 
           y = density   ), 
       data = redwine) +
      geom_point(alpha = 0.3, size = 1) +
      geom_smooth(method = "lm", se = FALSE,size=1)
```

Earlier I could have said that residual sugars is a throw away variable 
because it has only 0.0137 correlation to quality.  However, residual sugar 
can impact density which can negativitly impact alcohol 
which can impact wine quality.  I wonder if PCA would condense residual 
sugars, density, and alcohol together.  I also believe that removing the outliers
from residual sugar would change this chart significantly.


```{r echo=FALSE, Multivariate_Plots}
set.seed(415)
train<-sample_frac(redwine, 0.7)
sid<-as.numeric(rownames(train)) # because rownames() returns character
test<-redwine[-sid,]

forest <- randomForest(factor(quality_log) ~ volatile.acidity_log + alcohol_log 
                       + sulphates_log,
                              data=train, metric = "Accuracy")
print(forest) 


```

### OPTIONAL: Did you create any models with your dataset? Discuss the \
strengths and limitations of your model.

I chose the random forest (RF) to evaluate the dataset.  

Strengths of RF:
 + Limited parameter tuning ... default parameters often have great performance
 + Fast
 + Versitile
 + Great for feature selection because it evaluates a lot of 
 decision tree variations
 + Illustrates information gain by inclusion/exclusion of variables in the forest
 
Limitations of RF:
 + Poor Peformance: small data sets
 + Fragmentation Problem: each split has diminishing information gain
 + Prone to overfitting

```{r}
importance    <- importance(forest)
feature_importance <- data.frame(Variables = row.names(importance))

# Create a rank variable based on importance
rank <- feature_importance %>%
  mutate(Rank = paste0('#',dense_rank(desc(importance))))

# Use ggplot2 to visualize the relative importance of variables
ggplot(rank, aes(x = reorder(Variables, importance), 
    y = importance)) +
  geom_bar(stat='identity') + 
  geom_text(aes(x = Variables, y = 0.5, label = Rank),
    hjust=0, vjust=0.55, size = 4, colour = 'red') +
  labs(x = 'Variables') +
  coord_flip() + 
  theme_few()

ggcorr(redwine)

```

The random forest classifier is a powerful tool to predict the quality of wine.  

It's good to see this classifier confirms that alcohol (the feature with the 
highest correlation) is indeed the most important for predicting quality.

The correlation map is a great visual to show why we initially chose alcohol, 
volatile.acidity, and sulphates as the main features to explore.  
It also highlights the relationships we observed between other features 
such as density and residual sugars and alcohol.  I would use this to 
further explore additional features to test in the decision tree.
```{r}
log.rw <- log10(redwine + 1)

rw.pca <- prcomp(log.rw[,1:12],
                 center = TRUE,
                 scale. = TRUE)

autoplot(rw.pca, scale = 0, data = redwine)
```

PCA doesn't provide a clear view on components, so further analysis is needed to 
determine how features would be condensed into components.  However, it appears
from this graph that there could be two components (0,0) and (0,1)

------

# Final Plots and Summary



### Plot One
```{r echo=FALSE, Plot_One}
ggplot(redwine, aes(x = alcohol_log)) +
  geom_histogram(binwidth = 0.01) +
  labs(y = 'Alcohol %', x = 'Count',
          title = "Alcohol Log Scaled") 

```


### Description One

This graph is a great illustration that shows the impact of feature scaling alcohol.
The initial histogram of alcohol was very spread out and difficult to read, but
log transforming the data gives us a much cleaner look at this feature.

### Plot Two
```{r echo=FALSE, Plot_Two}
ggplot(aes(x = residual.sugar, y = density), data= redwine)+
  scale_x_continuous(lim = c(1, quantile(redwine$residual.sugar,0.90)))+
  scale_y_continuous(lim = c(0.99, quantile(redwine$density,0.90)))+    
  geom_point(fill = I('#F79420'), color = I('black'), shape =21) +
  stat_smooth(method = 'lm') +
    labs(y = 'Density g / cm^3', x = 'Residual Sugar g / dm^3',
          title = "Residual Sugar vs Density (Outliers Removed)")

```

### Description Two

I chose the graph above because I wanted to further investigate the relationship
between residual sugars and density.  There does still appear to be a direct linear
relationship between these variables even after removing outliers.  

### Plot Three
```{r echo=FALSE, Plot_Three}
log.rw <- log10(redwine + 1)

rw.pca <- prcomp(log.rw[,1:12],
                 center = TRUE,
                 scale. = TRUE)
autoplot(rw.pca, scale = 0, data = redwine,
         loadings = TRUE, loadings.colour = 'blue',
         loadings.label = TRUE, loadings.label.size = 5, alpha = 0.1) +
    labs(title = "PCA Components Graph by Feature")

```

### Description Three

I added the feature names to my PCA graph and found there were 4 components.
I found it fascinating that alcohol, sulphates and volatile acidity all belonged
to seperate components.  This makes me believe that these features selected 
are fairly balanced.  However, PH belongs to a seperate component and 
perhaps adding this feature would have increased the performance of the classifier.

------

# Reflection
I analyzed the data and built a model that could correclty predict 
the wine quality with a high accuracy based on alcohol content, sulphates, 
and volatile acidity.  

I do believe that accuracy can be further improved 
because the 3 features of interest only explained 35% of the variance, 
so my next step would be to run the model with the 4 PCA components 
that we discovered in final plot 3.  Feature selection and feature engineering
are areas that I struggled with. I took a trial and error approach by only 
selecting the 3 highest features with a correlation to quality.
However, an additional next step would be to use a pipeline of PCA, 
SelectKBest, and the random forest classifier to put together a more 
scientific approach to feature selection.

Overall, I am happy to have successfully built a base level starting point upon
which more indepth analysis can be performed.



